Minimal trap-spaces of Logical models are maximal siphons of their Petri net encoding

### 2022-05-02 cmsb22.tex
Mostly superficial stuff…

\documentclass[runningheads]{llncs}
\usepackage[T5]{fontenc}
% \usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}

@@

Given a Boolean model \(\mathcal{M} = (V, F)\) with variables \(V=(v_{1},\dots,v_{n})\) and Boolean functions \(F=(f_{1},\dots,f_{n})\), its state-space is \(\mathcal{S} =\bbbb^{n}\) with \(\bbbb = \{0, 1\}\).

A \emph{partial state} \(P\) of \(\mathcal{M}\) is a subset of \(\mathcal{S}\) where for some \(U\subset V\) the values are fixed by a mapping \(p: U\mapsto\bbbb\) and the other variables (\(V\setminus U\)) are \emph{free}.

  A \emph{trap space} of \(\mathcal{M}\) is a partial state \(P\) that is also a \emph{trap set}:

  if \(P'\) is the image of \(P\) by the state transition function, then \(P'\subset P\).

Here and in the following, the notion of \emph{minimality/maximality} will always be w.r.t. to subset inclusion.

%%% Subspace %%%
A subspace \(m\) of \(S_{\mathcal{M}}\) is characterized by its fixed variables (denoted by \(D_m\)) and free variables. This subspace can be specified by an assignment \(m : D_m \mapsto \bbbb\) where \(D_m \subseteq V\) and \(m(u)\) is the value of node \(u \in D_m\). The remaining variables of \(\mathcal{M}\), \(V \setminus D_m\), are said to be free, i.e., they can receive any Boolean value. We specify subspaces like states but use in addition the symbol \(\star\) to indicate that a variable is free. In addition, a subspace \(m\) also refers to the set of states \(\mathcal{S}_{\mathcal{M}}[m] := \{s \in \mathcal{S}_{\mathcal{M}}\;|\;\forall x_i \in D_m : s_i = m(x_i)\}\). For example, \(m = \star\star1\) means that \(D_m = \{x_3\}\), \(m(x_3) = 1\), and refers to the set of states \(\{001, 011, 101, 111\}\). Let \(\mathcal{S}_{\mathcal{M}}^{\star}\) denote the set of all possible subspaces of \(\mathcal{M}\). Note that \(\left|\mathcal{S}_{\mathcal{M}}^{\star}\right| = 3^n\) and \(S_{\mathcal{M}} \subset \mathcal{S}_{\mathcal{M}}^{\star}\)~\cite{klarner2015computing}.

@@

@@

Since our study is focused on Boolean models, we briefly recall the original encoding here.
Its basis is that every \emph{gene} of the original model is represented by two separate places, corresponding to its two states, active, and inactive.
Each conjunct of the logical function that activate the \emph{gene} will lead to a transition, consuming the inactive place, producing the active place, and with all other literals both consumed and produced.
And conversely for the inactivation.

@@

\section{Evaluation}
\subsection{Benchmark vs.\ implicant-based methods}

Using BioLQM~\ref{sec:encoding}

\url{http://colomoto.org/biolqm/doc/tools-trapspace.html}

@@

@@

% \begin{table}
% \caption{Table captions should be placed above the
% tables.}\label{tab1}
% \begin{tabular}{|l|l|l|}
% \hline
% Heading level &  Example & Font size and style\\
% \hline
% Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
% 1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
% 2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
% 3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
% 4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
% \hline
% \end{tabular}
% \end{table}

>>>>

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
% \usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{booktabs}

@@

Given a Boolean model \(\mathcal{M} = (V, F)\) with variables \(V=(v_{1},\dots,v_{n})\) and Boolean functions \(F=(f_{1},\dots,f_{n})\), its state-space is \(\mathcal{S}_{\mathcal{M}} =\bbbb^{n}\) with \(\bbbb = \{0, 1\}\).

%%% Subspace %%%
A \emph{subspace} \(m\) of \(\mathcal{S}_{\mathcal{M}}\) is characterized by its fixed variables (denoted by \(D_m\)) and free variables.
This subspace can be specified by an assignment \(D_m \mapsto \bbbb\), also denoted by \(m\) by abuse of notation, and where \(D_m \subseteq V\) and \(m(u)\) is the value of node \(u \in D_m\).
The remaining variables of \(\mathcal{M}\), \(V \setminus D_m\), are said to be \emph{free}, i.e., they can receive any Boolean value.
We write subspaces like states but use in addition the symbol \(\star\) to indicate that a variable is free.
A subspace \(m\) thus corresponds to the set of states \(\mathcal{S}_{\mathcal{M}}[m] := \{s \in \mathcal{S}_{\mathcal{M}}\;|\;\forall x_i \in D_m : s_i = m(x_i)\}\).
For example, \(m = \star\star1\) means that \(D_m = \{x_3\}\), \(m(x_3) = 1\), and corresponds to the set of states \(\{001, 011, 101, 111\}\).
Let \(\mathcal{S}_{\mathcal{M}}^{\star}\) denote the set of all possible subspaces of \(\mathcal{M}\). Note that \(\left|\mathcal{S}_{\mathcal{M}}^{\star}\right| = 3^n\) and \(S_{\mathcal{M}} \subset \mathcal{S}_{\mathcal{M}}^{\star}\)~\cite{klarner2015computing}.

@@

\sylvain{missing example?}

@@

Since our study is focused on Boolean models, we briefly recall the original encoding here.
Its basis is that every \emph{gene} of the original model is represented by two separate places (\(p\) and \(\overline{p}\)), corresponding to its two states, active, and inactive.
Each conjunct of the logical function that activate the \emph{gene} will lead to a transition, consuming the inactive place, producing the active place, and with all other literals both consumed and produced.
And conversely for the inactivation.

@@

\section{Evaluation}
To assess the efficiency of the proposed method, we compared it with the state-of-the-art method PyBoolNet~\cite{klarner2017pyboolnet}.
We used a set of real-world Boolean models lying in various scales obtained from the literature.
These models are quite big (in size), complex (i.e., having high average in-degree, which is related to the number of prime-implicants) and most of them have never been fully analyzed.
We then applied the proposed method and PyBoolNet to computing minimal trap spaces of these real-world models.
To solve the ASP problems, we used the same ASP solver CLINGO~\cite{DBLP:journals/aicom/GebserKKOSS11} and the same configuration as that used in PyBoolNet~\cite{klarner2015computing}.
Specifically, ...
Finally, we set a time limit as one hour for each model.

\sylvain{we need a source for the models as much as possible, or to refer to some place where people can access them to reproduce the benchmark.}

\subsection{Benchmark vs.\ implicant-based methods}

Table~\ref{tab:result_real} shows the experimental results on real-world models. Hereafter, we analyze in detail the results with respect to minimal trap space computation.

\begin{table}[!ht]
  \caption{Timing comparisons between the proposed method and PyBoolNet on the selected real-world models.
    Columns "\(n\)", "\(e\)", and "\(s\)" denote the numbers of nodes, interactions, and source nodes of each model, respectively.
    Column "\(|M|\)" denotes the number of minimal trap spaces.
    The computation times are all in seconds.
    "DNF" means that the method did not finish the computation within one hour.}
  \label{tab:result_real}
  \begin{tabular}{lrrrrrrrrr}
    \toprule
    & & & & & & \multicolumn{2}{c}{PyBoolNet} & \multicolumn{2}{c}{Proposed method}\\
    \cmidrule(lr){7-8}\cmidrule(lr){9-10}
    model & $n$ & $e$ & $s$ & $|M|$ & $|F|$ & t1 (s) & t2 (s) & t1 (s) & t2 (s) \\ \midrule
    CASCADE-3 & 183 & - & 0 & 1 & 0 & - & - & 1.65 & - \\
    CHO-2016 & 200 & - & 13 & 1024 & 0 & - & - & 2.10 & - \\
    CACC & 70 & 153 & 1 & 10 & 2 & - & - & 0.18 & - \\
    THIL\_17 & 92 & - & 16 & 5724 & 5724 & - & - & 0.90 & - \\
    ERBB & 247 & 1100 & 22 & 1456394+ & 0 & DNF & DNF & DNF & - \\
    ER-STRESS & 182 & 266 & 75 & 25600 & 25600 & - & - & 4.54 & - \\
    AD & 1169 & - & 469 & 49152 & 49152 & - & - & 66.17 & - \\
    Cholocystokinin & 404 & - & 95 & 705024 & 705024 & - & - & 1080.35 & - \\
    FIBROBLASTS & 139 & 557 & 9 & 375850 & 32768 & - & - & 331.87 & - \\
    TcellCheckPoint & 218 & - & 34 & 15360 & 10240 & - & - & 4.33 & - \\
    HIV-1 & 138 & 368 & 14 & 1408 & 1408 & DNF & DNF & 1.44 & - \\
    HMOX1 & 145 & 228 & 56 & 5120 & 5120 & - & - & 0.99 & - \\
    Human network & 1953 & - & 675 & 2048 & 2048 & DNF & DNF & 1770.98 & - \\
    IL\_6 Signalling & 86 & 149 & 15 & 512 & 448 & - & - & 0.32 & - \\
    IMMUNE-SYS & 164 & 506 & 13 & 36852 & 27871 & DNF & DNF & 14.19 & - \\
    IBD & 47 & 287 & 0 & 1 & 0 & DNF & DNF & 296.23 & - \\
    Influenza A & 131 & 302 & 11 & 2560 & 2560 & - & - & 1.07 & - \\
    INTERFERON-1 & 121 & 190 & 55 & 512 & 512 & - & - & 0.43 & - \\
    KORKUT-2015 & 99 & - & 12 & 2060 & 1384 & DNF & DNF & 2.86 & - \\
    KY-PA & 150 & 304 & 72 & 512 & 512 & - & - & 3.79 & - \\
    RA Apoptosis & 431 & - & 59 & 640 & 0 & DNF & DNF & 0.70 & - \\
    CELL CYCLE 2020 & 121 & 493 & - & 78 & 78 & - & - & 3.72 & - \\
    S1\_Table & 1659 & - & 522 & 1024 & 1024 & DNF & DNF & 1511.07 & - \\
    Macrophage\_Act & 321 & 533 & 19 & 512 & 384 & - & - & 1.58 & - \\
    SIPC & 133 & 431 & 0 & 812 & 746 & - & - & 1.40 & - \\
    SPBP & 66 & 139 & 13 & 512 & 256 & - & - & 0.17 & - \\
    SN5 & 2746 & - & 835 & 2048 & 2048 & DNF & DNF & 3094.04 & - \\
    TCR-TLR5-SIG & 130 & 273 & 2 & 48 & 48 & - & - & 0.50 & - \\
    TLGLSurvival & 61 & 193 & 7 & 318 & 172 & - & - & 0.24 & - \\
    TUREI-2016 & 4691 & - & 1257 & - & - & DNF & DNF & DNF & DNF \\
    VIRUS-RE-CYC & 154 & 268 & 25 & 1024 & 1024 & - & - & 0.67 & - \\ \bottomrule
  \end{tabular}

\sylvain{I changed the ``look'' of the table using booktabs, see \url{https://nhigham.com/2019/11/19/better-latex-tables-with-booktabs/} for instance.}

\sylvain{what is $|F|$? what are t1 and t2? why are there no times for BoolNet?}
\end{table}

@@

Also to a BDD-based approach using BioLQM is proposed at \url{http://colomoto.org/biolqm/doc/tools-trapspace.html} but because of the BDD conversion, some models are not even loaded.
This is also an issue for using BioLQM as a translator/preprocessor and since we do not use BDDs we might rely on writing our own parser for SBML-Qual and such formats.

@@
